---
output: github_document
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/emilio-berti/assembly/workflows/R-CMD-check/badge.svg)](https://github.com/emilio-berti/assembly/actions)
<!-- badges: end -->


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 4,
  fig.height = 4,
  fig.align='center',
  dpi = 300,
  out.width = "50%"
)
```

# Example workflow

Following food web terminology, I talk here about _resources_ and _consumers_. In the case of plant-pollinator networks, this is analogous as replacing _resource_ with _plant_ and _consumers_ with _pollinator_. In this case, however, some filtering steps may be unnecessary and unwanted. __More about this in a separate section.__

I show an example workflow that is composed of the following steps:

1. draw random species from a metaweb
2. impose resource filtering, i.e. each basal species must be consumed and each consumer must feed on a resource.
3. impose limiting similarity filtering, i.e. consumers are constantly replaced by others following a probability distribution proportional to their similarity of interactions. 

## Draw random species from a metaweb

Loading the required libraries and set a random seed:

```{r init}
library(assembly)
library(igraph)

set.seed(1234)
```

Load the dataset _adirondack_ that comes with _assembly_:

```{r adirondack}
data(adirondack)
```

_adirondack_ is the Adirondack Lakes metaweb as obtained from the GATEWAy database.

Define the number of species for the local community:
```{r S}
S <- 50 #species richness
```

To draw a random community, use the function `draw_random_species()`:

```{r random}
sp <- draw_random_species(S, colnames(adirondack))
sum(colSums(adirondack[sp, sp]) == 0)
show_graph(sp, adirondack)
```

## Hidden functions

There are several hidden functions in _assembly_. The reason there are hidden functions is because there is no need to call them directly. Hidden functions can be accessed by prefixing the `assembly:::` (three colon, not two). All hidden functions start with a dot `.`, e.g. `assembly:::.basals()`.

In general, you should not be bothered by hidden functions and should not call them directly, unless you have a good understanding of how they operate. Nevertheless, I summarize them for clarity.

`assembly:::.basals()` get all basal species in the metaweb, and is equivalent to subset the names of the metaweb where `colSums(metaweb) == 0`:

```{r basals}
identical(
  sort(intersect(assembly:::.basals(adirondack), sp)),
  sort(intersect(colnames(adirondack)[colSums(adirondack) == 0], sp))
)
```

`assembly:::.consumers()` and `assembly:::.top()` return the consumers and top consumers of the metaweb, respectively.

`assembly:::.find_isolated()` returns the species that are isolated in the local community:

```{r find_isolated}
assembly:::.find_isolated(sp, adirondack)
```
`assembly:::.find_replacements()` find suitable replacement for the isolated species:

```{r find_replacements}
assembly:::.find_replacements(sp,
                              assembly:::.find_isolated(sp, adirondack),
                              adirondack,
                              keep.n.basal = TRUE)
```

If _keep.n.basal_ is TRUE (default = FALSE), then the original number of basal species will not change.

`assembly:::.move()` performs a move in the limiting similarity procedure (more about this later):

```{r move}
tryCatch(assembly:::.move(sp, adirondack, t = 1),
         error = function(e) print(e))
```

This call to `assembly:::.move()` fails because isolated species are detected in the input. This is a desired property of the function, i.e. it fails when there is an unexpected behavior. All hidden functions have some kind of behavior-check, which is a safety net to assure the code is doing what you asked for.

Finally, `assembly:::.components()` returns the number of connected components in the graph of the local community:

```{r components}
assembly:::.components(sp, adirondack)
```

Usually, a proper food web has only one component, i.e. all species are connected by a path. Having more than one component means that the food web is actually made of several disconnected communities. In the case above, it also means that at least one of this disconnected communities is composed of only one isolated species.

# Resouce filtering

To impose the resource filtering, I call the function `resource_filtering()`. This takes as input the species names, the metaweb, and an optional argument _keep.n.basal_ to specify weather the original number of basal species should be kept constant (default = `FALSE`). **NOTE this may not be implemented correctly**

Behind the curtain, `resource_filtering()` calls the hidden functions as a way to compress code and make it consistent. That's why you shouldn't bother too much about hidden functions: they're there because they're useful in the development of the package, rather than for your usage. If they're useful for you and you understand how they work, use them.

```{r resource}
sp_resource <- resource_filtering(sp, adirondack, keep.n.basal = TRUE)
show_fw(sp_resource, adirondack, title = "Resource filtering")
show_graph(sp_resource, adirondack)
```

Now the local community is fully connected, i.e. basal species always have a consumer and consumers always have an available resource. It's possible to check this manually calling the hidden functions and working on the adjacency matrix of the local community:

```{r check_connected}
bas <- intersect(sp_resource, assembly:::.basals(adirondack))
cons <- intersect(sp_resource, assembly:::.consumers(adirondack))
all(rowSums(adirondack[bas, cons]) > 0)
all(colSums(adirondack[union(bas, cons), cons]) > 0)
```

Usually you don't need to perform these checks, as I implemented them within  `resource_filtering()`. I also implemented a check for disconnected components, to make sure that the resulting community has no isolated species and only one actual community. 

Bonus: because of these checks, now it is safe to perform a move of the limiting similarity procedure:

```{r fail}
assembly:::.move(sp_resource, adirondack, t = 1)
```

# Limiting similarity filtering

The limiting similarity filtering is composed of a series of individuals moves:

  1. A metric ($J$) representing the similarity of interaction is calculated for each species in the community.
  2. One species is removed with probability proportional to their $J$.
  3. The species removed is replaced by another species selected at random from the metaweb.

Each move can then be accepted of discarded based on a probabilistic acceptance criterion, the Metropolis-Hasting algorithm (see below for details). If the move is accepted, the new community will differ from the starting one only in the removed/replaced species. If the move is discarded, the new community is identical to the starting one. Each move takes as input the community of the previous move (or the initial community for the first move) and returns as output the new community (or the original one). 

The limiting similarity procedure is simply a series of moves (accepted or not). The starting community for the limiting similarity procedure is usually a community that has undergone already a resource filtering, imposing thus that trophic competition comes after resource availability. However, it is possible to skip the resource filtering step; the only requirement of the starting community for the limiting similarity filtering is that it does not have isolated species and has only one connected components.

To impose the limiting similariy filtering, call the function `similarity_filtering()`. This function has the argument _t_, which is the temperature of the Metropolis-Hastings algorithm, and _max.iter_ (default = 1,000), which is the maximum number of moves allowed.

```{r limiting}
sp_sim <- similarity_filtering(sp_resource, adirondack, t = 1)
show_graph(sp_sim, adirondack)
```

## Metropolis-Hasting algorithm

To accept a move, it must pass a Metropolis-Hasting algorithm. If the similarity of the new community is lower than the similarity of the old one, the move is always accepted. When the new similarity is higher than the old similarity, the move is accepted if:

$$
e^{\left( 1 - \frac{similarity_{new}}{similarity_{old}} \right) \frac{1}{t}} > \mathcal{U}(0, 1)
$$
This means that, even when the new similarity is higher than the old one and the new community has species with increased similarity of interaction, it can still be accepted as a valid move based on a probability density function. The probability of acceptance depends on how much the new similarity is higher than the old one and by the temperature parameter _t_. For increasing _t_, it is more likely to accept a non-favorable move:

```{r metro, fig.height=5, fig.width=6, out.width="80%"}
temp <- 10 ^ seq(-2, 1, by = .1)
ratio <- 10 ^ seq(-1, 2, by = .1)

move <- rep(NA, length(temp) * length(ratio))
i <- 1
for (t in temp){
  for (x in ratio) {
    move[i] <- metropolis.hastings(1, x, t)
    i <- i + 1
  }
}

d <- data.frame(Temp = rep(temp, each = length(ratio)),
                Ratio = rep(ratio, length(temp)),
                Move = as.numeric(move))
cols <- colorRampPalette(c("steelblue", "tomato"))
cols <- cols(length(unique(d$Temp)))
pal <- adjustcolor(cols, alpha.f = .2)
pal <- pal[sapply(d$Temp, \(x) which(unique(d$Temp) == x))]
plot(log10(d$Ratio), jitter(d$Move, factor = .2), 
     xlab = "Similiarity ratio (new / old)",
     ylab = "Accepted move",
     col = pal, pch = 20, frame = FALSE)
for (t in unique(d$Temp)) {
  x <- log10(d$Ratio[d$Temp == t])
  y <- d$Move[d$Temp == t]
  fit <- loess(y ~ x)
  lines(fit$x, fit$fitted, col = cols[which(unique(d$Temp) == t)])
}
legend(-1, .7, legend = seq(-2, 1, by = .5), 
       fill = colorRampPalette(c("steelblue", "tomato"))(7),
       title = "log10(t)")
```

The reason to include this algorithm is to avoid a purely deterministic procedure and include some stochasticity in the process. However, if this is unwanted, it can be removed (and the process made purely deterministic), by specifying a very high temperature parameter _t_:

```{r tweak_t}
table(sapply(seq_len(1000), \(x) metropolis.hastings(1, 1e3, t = 1e9)))
```

## Example 1: Trophic levels in random and filtered communities

For 50 local communities I:

1. draw random species
2. apply the resource filtering procedure
3. calculate the trophic level of the species in the local communities
4. I calculate average and maximum trophic levels within each community

```{r trolev}
# TL_random <- matrix(NA, S, 50)
# TL_resource <- matrix(NA, S, 50)
# for (i in seq_len(50)) {
#   sp <- draw_random_species(S, colnames(adirondack))
#   sp_resource <- resource_filtering(sp, adirondack, keep.n.basal = TRUE)
#   TL_random[, i] <- ATNr::TroLev(adirondack[sp, sp])[, 1]
#   TL_resource[, i] <- ATNr::TroLev(adirondack[sp_resource, sp_resource])[, 1]
# }
# plot(colMeans(TL_random), colMeans(TL_resource),
#      xlim = c(1.1, 2.2), ylim = c(1.1, 2.2),
#      main = "Average trophic level",
#      xlab = "Random communities",
#      ylab = "Resource filtered communities")
# abline(0, 1)
# plot(apply(TL_random, 2, max), apply(TL_resource, 2, max),
#      xlim = c(2.2, 9.2), ylim = c(2.2, 9.2),
#      main = "Maximum trophic level",
#      xlab = "Random communities",
#      ylab = "Resource filtered communities")
# abline(0, 1)
```

Trophic level tend to be higher in resource-filtered communities compared to random communities. This is due to isolated consumers (trophic level = 1) in the random communities and that were replaced by connected consumers when imposing resource filtering. It's also evident when computing the number of connected components:

```{r}
show_fw(sp, adirondack)
assembly:::.components(sp, adirondack)

show_fw(sp_resource, adirondack)
assembly:::.components(sp_resource, adirondack)
```
