---
title: "assembly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{assembly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Example workflow

Following food web terminology, I talk here about _resources_ and _consumers_. In the case of plant-pollinator networks, this is analogous as replacing _resource_ with _plant_ and _consumers_ with _pollinator_. In this case, however, some filtering steps may be unnecessary and unwanted. __More about this in a separate section.__

I show an example workflow that is composed of the following steps:

  1. draw random species from a metaweb
  2. impose resource filtering, i.e. each basal species must be consumed and each consumer must feed on a resource.
  3. impose limiting similarity filtering, i.e. consumers are constantly replaced by others following a probability distribution proportional to their similarity of interactions. 

## Draw random species from a metaweb

```{r random, fig.align='center', fig.width=4, fig.height=4}
library(assembly)
library(igraph)

set.seed(1234)

# load Adirondack dataset
# this is part of the assembly package
data(adirondack) 

S <- 50 #species richness
sp <- draw_random_species(S, colnames(adirondack)) #draw 50 random species
sum(colSums(adirondack[sp, sp]) == 0) #20 basal species
show_fw(sp, adirondack, title = "Random")
plot(graph_from_adjacency_matrix(adirondack[sp, sp]), vertex.label = NA)
```

You can see that some species are isolated, i.e. they do not interact with any other species in the local community. You can find isolated species with the hidden function ```assembly:::.find_isolated()```.

```{r hidden_isolated}
assembly:::.find_isolated(sp, adirondack)
```
Similarly, it's possible to obtain the number of weakly connected components of the graph of the local community:

```{r components}
assembly:::.components(sp, adirondack)
```

The reason there are hidden functions in _assembly_ is because there is no need to call them directly. Hidden functions can be accessed by prefixing the `assembly:::` (three colon, not two).

To impose the resource filtering, I call the function `resource_filtering()`. This takes as input the species names, the metaweb, and an optional argument _keep.n.basal_ to specify weather the original number of basal species should be kept constant (default = `FALSE`). **NOTE this still not implemented correctly***

```{r resource, fig.align='center', fig.width=4, fig.height=4}
sp_resource <- resource_filtering(sp, adirondack, keep.n.basal = TRUE)
show_fw(sp_resource, adirondack, title = "Resource filtering")
plot(graph_from_adjacency_matrix(adirondack[sp_resource, sp_resource]), vertex.label = NA)
```
Now the local community is fully connected, i.e. basal species always have a consumer and consumers always have an available resource. It's possible to check this manually calling other hidden functions and working on the adjacency matrix of the local community:

```{r check_connected}
bas <- intersect(sp_resource, assembly:::.basals(adirondack))
cons <- intersect(sp_resource, assembly:::.consumers(adirondack))
all(rowSums(adirondack[bas, cons]) > 0)
all(colSums(adirondack[union(bas, cons), cons]) > 0)
```

In general, there is no need to check this, as if this is not the case `resource_filtering()` will fail: __put example here.__

```{r fail}
# nothing for now
```

## Example 1: Trophic levels in random and filtered communities

For 50 local communities I:

  1. draw random species
  2. apply the resource filtering procedure
  3. calculate the trophic level of the species in the local communities
  4. I calculate average and maximum trophic levels within each community

```{r trolev}
TL_random <- matrix(NA, S, 50)
TL_resource <- matrix(NA, S, 50)
for (i in seq_len(50)) {
  sp <- draw_random_species(S, colnames(adirondack))
  sp_resource <- resource_filtering(sp, adirondack, keep.n.basal = TRUE)
  TL_random[, i] <- ATNr::TroLev(adirondack[sp, sp])[, 1]
  TL_resource[, i] <- ATNr::TroLev(adirondack[sp_resource, sp_resource])[, 1]
}
plot(colMeans(TL_random), colMeans(TL_resource),
     xlim = c(1.1, 2.2), ylim = c(1.1, 2.2),
     main = "Average trophic level",
     xlab = "Random communities",
     ylab = "Resource filtered communities")
abline(0, 1)
plot(apply(TL_random, 2, max), apply(TL_resource, 2, max),
     xlim = c(2.2, 9.2), ylim = c(2.2, 9.2),
     main = "Maximum trophic level",
     xlab = "Random communities",
     ylab = "Resource filtered communities")
abline(0, 1)
```

Trophic level tend to be higher in resource-filtered communities compared to random communities. This is due to isolated consumers (trophic level = 1) in the random communities and that were replaced by connected consumers when imposing resource filtering. It's also evident when computing the number of connected components:

```{r}
show_fw(sp, adirondack)
assembly:::.components(sp, adirondack)

show_fw(sp_resource, adirondack)
assembly:::.components(sp_resource, adirondack)
```
